Capturing the shape of our experiences and how we remember them

### 2018-08-21 paper/main.tex
blur figure 1, edits to supp materials

\begin{figure}[th!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participant's verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in orange) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

@@

An important question for future work concerns the factors that drive an individual to sample their recall trajectory finely or coarsely. For example, given a short recall interval, would participants intuitively gravitate towards coarser samplings that still outline the basic shape of the movie's topic trajectory? Or if participants were told that their narrations would be played back to other participants ~\citep{ZadbEtal16}, would that change the resolution or shape of their recalls? And over successive recountings of the same sequence of events or with more elapsed time between encoding and recall, how do the shapes of the trajectories change? For example, loss of detail would result in a "smoothing out" of the trajectories with each new retelling.

@@

\section{Additional measures of naturalistic memory}
To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered and computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p$<.05, permutation test, see Methods for details; participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to hand annotated memory performance across participants (Pearson's $r$(16)=-.57, $p$=.016). Thus, a closer match between the video and recall event models was related to better recall performance.

Next, we tested whether participants who recalled more events were also more precise in their recollections. For each participant, we computed the correlation between each recall event and its matching video event (only for the events which they recalled). This resulted in a single number for each recalled event indexing how similar the recall event was to its matching movie event (i.e the ``precision'' of the recall). We then averaged the correlations within participant. In line with our prediction, there was a strong correlation between hand annotated memory performance and precision suggesting that participants who remembered more events also remembered them more veridically (Pearson's $r$(16)=.74, $p$=.0006). Then, we considered the distinctiveness of each recall event. That is, how uniquely a recall event matched a given video event compared to all other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way). To this end, we computed a `distinctiveness' score for each participant (i.e. 1 - the correlation between a recall event and all non-matching video events).  Then, we averaged this measure over recall events within participant.  We found that participants with higher hand annotated memory performance also had higher distinctiveness scores (Pearson's $r$(16)=.8, $p$=.0001).

@@

\section{Naturalistic extensions of classic list-learning analyses}
Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. In our first set of analyses, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings. First, we asked whether the estimated number of recall events ($k$) by participant was related to hand-annotated accuracy as published in Chen et al. (2017).  We found a strong positive correlation where subjects with a greater number of recall events also had better overall memory performance (Pearson's $r$(16)=.67, $p$=.003). Then, we considered how participants initiated the recall sequence (known in the literature as the `probability of first recall' or `PFR'). We found that participants tended to initiate their recall sequences with the first few events (Supp. Fig.~\ref{fig:list-learning}a), which is qualitatively very similar to previously published list learning experiments (REF?). Next, we considered another well-studied memory measure in the list-learning literature, the lag conditional response probability curve (or lag-CRP) ~\citep{Kaha96}. The result suggests a strong bias to transition sequentially events in the forward direction (Supp. Fig.~\ref{fig:list-learning}b). Finally, we assessed memory performance for each event in the video as a function of its serial position during encoding (Supp. Fig.~\ref{fig:list-learning}c). We also considered two additional across-participant measures of recall that characterize memory organization: temporal clustering and semantic clustering. We found that participants who clustered in time also recalled a greater number of events (Pearson's $r$(16)=.62, $p$=.007). Next, we assessed semantic clustering. We found that the semantic clustering score was related to memory performance across participants (Pearson's $r$(16)=.55, $p$=.02).  Thus, participants who organized their recalls with respect to the semantic information contained in the scene had better memory performance.

@@

\section{Supplemental Methods}

\subsection{Characterizing memory performance using traditional approaches}
\textbf{Overall Accuracy.}  To get an overall measure of the quantity of information recalled, we computed the proportion of sucessfully recalled events by counting the number of unique recall events identified by the HMM model and dividing by the total number of video events.  We performed this analysis for each participant separately.

@@

>>>>

\begin{figure}[th!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic_blur.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participant's verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in orange) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

@@

An important question for future work concerns the factors that drive an individual to sample their recall trajectory finely or coarsely. For example, given a short recall interval, would participants intuitively gravitate towards coarser samplings that still outline the basic shape of the movie's topic trajectory? Or if participants were told that their narrations would be played back to other participants ~\citep{ZadbEtal17}, would that change the resolution or shape of their recalls? And over successive recountings of the same sequence of events or with more elapsed time between encoding and recall, how do the shapes of the trajectories change? For example, loss of detail would result in a "smoothing out" of the trajectories with each new retelling.

@@

\section{Naturalistic extensions of classic list-learning analyses}
Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. Here, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings. First, we asked whether the estimated number of recall events ($k$) by participant was related to hand-annotated accuracy as published in Chen et al. (2017).  We found a strong positive correlation where subjects with a greater number of recall events also had better overall memory performance (Pearson's $r$(16)=.67, $p$=.003). Then, we considered how participants initiated the recall sequence (known in the literature as the `probability of first recall' or `PFR'). We found that participants tended to initiate their recall sequences with the first few events (Supp. Fig.~\ref{fig:list-learning}a), which is qualitatively very similar to previously published list learning experiments ~\citep{HowaKaha99}. Next, we considered another well-studied memory measure in the list-learning literature, the lag conditional response probability curve (or lag-CRP) ~\citep{Kaha96}. The result suggests a strong bias to transition sequentially events in the forward direction (Supp. Fig.~\ref{fig:list-learning}b). Finally, we assessed memory performance for each event in the video as a function of its serial position during encoding (Supp. Fig.~\ref{fig:list-learning}c). We did not observe the classic ``primacy'' and ``recency'' pattern which is prevalent in the literature ~\citep{Murd62a}. We also considered two additional across-participant measures of recall that characterize memory organization: temporal clustering and semantic clustering. We found that participants who clustered in time also recalled a greater number of events (Pearson's $r$(16)=.62, $p$=.007). Next, we assessed semantic clustering. We found that the semantic clustering score was related to memory performance across participants (Pearson's $r$(16)=.55, $p$=.02).  Thus, participants who organized their recalls with respect to the semantic information contained in the scene had better memory performance.

\section{Additional measures of naturalistic memory}
To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered and computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p$<.05, participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to hand annotated memory performance across participants (Pearson's $r$(16)=-.57, $p$=.016). Thus, a closer match between the video and recall event models was related to better recall performance.

Next, we tested whether participants who recalled more events were also more precise in their recollections. For each participant, we computed the correlation between each recall event and its matching video event (only for the events which they recalled). This resulted in a single number for each recalled event indexing how similar the recall event was to its matching movie event (i.e the ``precision'' of the recall). We then averaged the correlations within participant. In line with our prediction, there was a strong correlation between hand annotated memory performance and precision suggesting that participants who remembered more events also remembered them more veridically (Pearson's $r$(16)=.74, $p$=.0006).

Then, we considered the distinctiveness of each recall event. That is, how uniquely a recall event matched a given video event compared to all other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way). To this end, we computed a `distinctiveness' score for each participant (i.e. 1 - the correlation between a recall event and all non-matching video events).  Then, we averaged this measure over recall events within participant.  We found that participants with higher hand annotated memory performance also had higher distinctiveness scores (Pearson's $r$(16)=.8, $p$=.0001).

@@

@@

\section{Supplemental Methods}
\subsection{List-learning analyses}
\textbf{Overall Accuracy.}  To get an overall measure of the quantity of information recalled, we computed the proportion of sucessfully recalled events by counting the number of unique recall events identified by the HMM model and dividing by the total number of video events.  We performed this analysis for each participant separately.

@@

\subsection{Additional measures of naturalistic memory}
\textbf{Precision.} This measure gives us an indication of the specific match between a video event and recall event, where values approaching 1 are highly precise and lower values are inprecise.  We defined ``precision'' as the correlation between a recall event and its matching (i.e. argmax) video event.

\textbf{Distinctiveness.} Distinctiveness quantifies how similar a recall event is to all non-matching video events.  It provides a metric of how uniquely a particular recall event describes a particular video event.  To compute it, a given recall event is correlated to all video events, the argmax is removed and the rest of the values are averaged. The resulting value is subtracted from 1 such that larger values indicate a more distinctive recall event.
